---
title: "ST 558 Project 2"
author: "David Arthur"
date: "6/28/2021"
output:
  github_document:
    toc: true
always_allow_html: true
params:
  dayOfWeek: "Monday"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(caret)
library(corrplot)
library(ggplot2)
library(GGally)
library(knitr)
library(faraway)
library(leaps)
```


Read in data, filter by day of week
```{r message = FALSE}
day <- readr::read_csv("day.csv", col_types = cols(
  season = col_factor(),
  yr = col_factor(),
  mnth = col_factor(),
  holiday = col_factor(),
  weekday = col_factor(),
  workingday = col_factor(),
  weathersit = col_factor()))

day <- day %>% mutate(season = fct_recode(season, winter = "1", spring = "2", summer = "3", fall = "4")) %>%
  mutate(yr = fct_recode(yr, "2011" = "0", "2012" = "1")) %>%
  mutate(weekday = fct_recode(weekday, Sunday = "0", Monday = "1", Tuesday = "2", Wednesday = "3", Thursday = "4", Friday = "5", Saturday = "6")) %>%
  mutate(weathersit = fct_recode(weathersit, clear = "1", mist = "2", lightRainOrSnow = "3", heavyRainOrSnow = "4")) %>%
  filter(weekday == params$dayOfWeek)

dayNF <- readr::read_csv("day.csv")
```

Partition data into training and test sets
```{r}
set.seed(21)
trainIndex <- createDataPartition(day$cnt, p = 0.7, list = FALSE)
dayTrain <- day[trainIndex, ]
dayTest <- day[-trainIndex, ]
```

Exploratory data analysis and summary (David)
```{r message = FALSE}
GGally::ggpairs(dayTrain %>% select(3:9, atemp, windspeed, casual, registered, cnt))
# dayNFCor <- cor(as.matrix(dayNF %>% select(3:9, atemp, windspeed, casual, registered,cnt)))
# corrplot(dayNFCor, type = "upper", tl.pos = "lt")
# corrplot(dayNFCor, type = "lower", method = "number", add = TRUE, diag = FALSE, tl.pos = "n")
```

Exploration of individual predictors
```{r}
g <- ggplot(data = dayTrain)
g + geom_point(aes(x = dteday, y = registered))

meanByMonth <- dayTrain %>% group_by(mnth) %>%
  summarize(meanCas = mean(casual), meanReg = mean(registered), meanTotal = mean(cnt))
g2 <- ggplot(meanByMonth, aes(x = mnth))
g2 + geom_bar(aes(y = meanCas), stat = "identity")
g2 + geom_bar(aes(y = meanReg), stat = "identity")
g2 + geom_bar(aes(y = meanTotal), stat = "identity")

meanByYear <- dayTrain %>% group_by(yr) %>%
  summarize(meanCas = mean(casual), meanReg = mean(registered), meanTotal = mean(cnt))
g2 <- ggplot(meanByYear, aes(x = yr))
g2 + geom_bar(aes(y = meanCas), stat = "identity")
g2 + geom_bar(aes(y = meanReg), stat = "identity")
g2 + geom_bar(aes(y = meanTotal), stat = "identity")

meanByWeather <- dayTrain %>% group_by(weathersit) %>%
  summarize(meanCas = mean(casual), meanReg = mean(registered), meanTotal = mean(cnt))
kable(meanByWeather, digits = 1, col.names = c("Weather", "Mean Casual Riders", "Mean Registered Riders", "Mean Total Riders"), caption = "Average # of riders by weather category")

meanByHoliday <- dayTrain %>% filter(workingday == 0) %>%
  group_by(holiday) %>%
  summarize(meanCas = mean(casual), meanReg = mean(registered), meanTotal = mean(cnt))
kable(meanByHoliday, digits = 1, col.names = c("Holiday (0 = no, 1 = yes)", "Mean Casual Riders", "Mean Registered Riders", "Mean Total Riders"), caption = "Average # of riders on holidays vs. non-holiday non-workdays")

```

Exploratory data analysis and summary (James)
```{r carr_explore, message=FALSE, fig.height = 9, fig.width = 15}
ggpairs(dayTrain %>% select(-instant,-dteday, -season, -yr, -cnt), 
        ggplot2::aes(colour=workingday))

```
Notes from looking at ggpairs plots:
Working days are the highest usage for registered riders, but non-working days are the highest usage for casual riders. Registered riders are the primary volume, so we definitely care most about them but worth keeping in mind. There are two types of non-working days: weekends and holidays, and there is a difference in volume for each of those rider types depending on whether it is a holiday or a weekend. 

Air temperature and temperature are nearly 100% correlated. We should probably figure out which one of them we want to use. Speaking of correlated, can we drop the date and only use months? Unfortunately, it looks like we need to keep the year field as well, since year 2 had better performance than year 1. Do we want to keep season and month? I lean towards keeping year and month, but dropping season and date. Let me know what you think.

Looking at the scatter of casual vs registered, broken out by working day, it's crazy how separate the linear relationships look:
```{r}
g <- ggplot(data=dayTrain, aes(x=registered, y=casual))
g + geom_point(aes(color=workingday))

```
On working days, registered bikes are the main rider group. On non-working days, it switches to casual. Looking at day of the week, we may be able to exclude it since it will be covered by the working day flag and holiday flag, but I guess we can check the models to see if it provides anything extra.

```{r}
g <- ggplot(data=dayTrain %>% 
                 select(weekday, casual, registered) %>%
                 pivot_longer(cols=c(casual, registered),
                              names_to = 'metrics',
                              values_to = 'riders') %>%
                 group_by(weekday, metrics) %>%
                 summarise(avg_riders = mean(riders)), 
            aes(x=weekday, y=avg_riders, fill=metrics))
g + geom_bar(stat='identity', position='dodge')
```
Looking at this graph, weekday definitely seems relatively stable across the days (working days for registered and non-working days for casual are the jumps), but there may be enough variation to include it. 

##I like this graph.  I thought about doing one like it, but wasn't sure how to code it.  pivot_longer hadn't occurred to me.

##About which variables to include, I agree with your comments.  My understanding is that each of us comes up with our own models (I do a linear regression and a random forest, you do a linear regression and a boosted tree), so you and I don't need to include the same predictors.  We do need to agree ahead of time on which response we're going to model (casual, registered, or cnt), so that the results of the 4 models can be compared to each other.  I'm fine with any of the 3.  Do you have a preference?

Yeah, no preference here either. I guess we could just say registered since it's the highest volume customer, and if we were doing this analysis for that company then registered users would be the most important group.

##Sounds good, we'll go with registered.


```{r carr_bestsub, error = TRUE}
library(leaps)

data <- dayTrain %>% 
               filter(weekday == params$dayOfWeek) %>% drop_na() %>%
               select(-instant,-dteday, -season, 
                    -weekday, -atemp, -casual, -cnt)

#this function converts new data to a model matrix
#so that a prediction can be run via matrix multiplication
#on a best subsets model
predict.regsubsets = function(object,newdata,id,...){
      form = as.formula(object$call[[2]]) 
      mat = model.matrix(form,newdata)    
      coefi = coef(object,id=id)          
      xvars = names(coefi)                
      mat[,xvars]%*%coefi               
}


#let's do cross validation with folds
k <- 4
set.seed(21)
folds <- sample(1:k, nrow(data), replace=T)

cv_errors = matrix(NA, k, 16, dimnames = list(NULL, paste(1:16)))

for (j in 1:k) {
  best <- regsubsets(registered ~ ., 
                     data=data[folds!=j,], nvmax=20)
  
  for (i in 1:16) {
    pred <- predict(best, data[folds==j,], id=i)
    
    
    cv_errors[j, i] <- mean((temp_data$registered[folds==j]-pred)^2)
  }
}

# Take the mean of over all folds for each model size
mean_cv_errors = apply(cv_errors, 2, mean)

# Find the model size with the smallest cross-validation error
min = which.min(mean_cv_errors)

#the model w/ 14 variables was best when using 4 fold cv.
#i did 4 fold because there are only about 80 rows of data per weekday

best_full <- regsubsets(registered ~ ., 
                     data=temp_data[folds!=j,], nvmax=20)

```

```{r}
fit <- lm(registered ~ temp*hum,
        data=dayTrain %>% 
             filter(weekday == params$dayOfWeek) %>% drop_na() %>%
             select(-instant,-dteday, -season, 
                    -weekday, -atemp, -casual, -cnt))
```

```{r}
names(dayTrain)
```


```{r arthur_bestMLR, message = FALSE, warning = FALSE}
# GGally::ggpairs(dayTrain %>% select(3:9, atemp, windspeed, casual, registered, cnt))
dayNFCor <- cor(as.matrix(dayNF %>% select(3:13, registered)))
corrplot(dayNFCor, type = "upper", tl.pos = "lt")
corrplot(dayNFCor, type = "lower", method = "number", add = TRUE, diag = FALSE, tl.pos = "n")

mlrFull <- lm(registered ~ dteday + season +  yr + mnth + holiday + workingday + weathersit + temp + atemp + hum + windspeed, dayTrain)
summary(mlrFull)
AIC(mlrFull)
x <- model.matrix(mlrFull)[, -1]
e <- eigen(t(x) %*% x)
e$val
sqrt(e$val[1]/min(e$val))
vif(x)

#remove dteday (high collinearity, low p-value)
mlr2 <- update(mlrFull, . ~ . - dteday)
summary(mlr2)
AIC(mlr2)
x <- model.matrix(mlr2)[, -1]
e <- eigen(t(x) %*% x)
e$val
sqrt(e$val[1]/min(e$val))
vif(x)
#reduced AIC and condition number

#remove temp (high collinearity, low p-value)
mlr3 <- update(mlr2, . ~ . - temp)
summary(mlr3)
AIC(mlr3)
x <- model.matrix(mlr3)[, -1]
e <- eigen(t(x) %*% x)
e$val
sqrt(e$val[1]/min(e$val))
vif(x)
#reduced AIC and condition number

#remove workingday (high collinearity, low p-value)
mlr4 <- update(mlr3, . ~ . - workingday)
summary(mlr4)
AIC(mlr4)
x <- model.matrix(mlr4)[, -1]
e <- eigen(t(x) %*% x)
e$val
sqrt(e$val[1]/min(e$val))
vif(x)
#reduced AIC and condition number

#remove season (high collinearity, low p-value)
mlr5 <- update(mlr4, . ~ . - season)
summary(mlr5)
AIC(mlr5)
#don't remove season because AIC increased

#remove month (high collinearity, low p-value)
mlr6 <- update(mlr4, . ~ . - mnth)
summary(mlr6)
AIC(mlr6)
x <- model.matrix(mlr6)[, -1]
e <- eigen(t(x) %*% x)
e$val
sqrt(e$val[1]/min(e$val))
vif(x)
#reduced AIC and condition number

#remove year from full model instead of dteday (high collinearity)
mlr7 <- update(mlrFull, . ~ . - yr - workingday - temp, dayTrain)
summary(mlr7)
AIC(mlr7)
x <- model.matrix(mlr7)[, -1]
e <- eigen(t(x) %*% x)
e$val
sqrt(e$val[1]/min(e$val))
vif(x)
#higher AIC than mlr4

#compare to model chosen by leaps::step() function
mlrStep <- step(mlrFull)
names(mlrStep)
mlrStep$call
mlr6$call
AIC(mlr6, mlrStep)
#my choice agrees with step() choice

#diagnostics plot
plot(mlr6$fitted, mlr6$residuals)
#indication of mild nonconstant variance
MASS::boxcox(mlr6)
#Box-Cox lambda close to 1, so no need for transformation of response

#look for nonlinearity with partial residuals plots
termplot(mlr6, partial.resid = TRUE)
#atemp, hum, and windspeed look somewhat nonlinear, so try quadratic terms for them
mlr8 <- update(mlr6, . ~ . + I(atemp^2))
summary(mlr8)
AIC(mlr8)
#improved AIC, so keep atemp^2 in model

mlr9 <- update(mlr8, . ~ . + I(hum^2))
summary(mlr9)
AIC(mlr9)
#improved AIC, so keep hum^2 in model

mlr10 <- update(mlr9, . ~ . + I(windspeed^2))
summary(mlr10)
AIC(mlr10)
#slightly improved AIC, compare using cross validation

#interaction of weather vars w/ holiday seems possible, so try adding to model
mlr11 <- update(mlr9, . ~ . + weathersit:holiday)
summary(mlr11)
AIC(mlr11)
#slightly worse AIC, compare using cross validation
mlr12 <- update(mlr9, . ~ . + atemp:holiday)
summary(mlr12)
AIC(mlr12)
#marginal decrease in AIC, compare using cross validation

#fit best candidate models using cross validation w/ caret package
mlrFit9 <- train(registered ~ season + yr + holiday + weathersit + atemp + hum + windspeed + I(atemp^2) + I(hum^2), data = dayTrain,
    method = "lm",
    preProcess = c("center", "scale"),
    trControl = trainControl(method = "repeatedcv", number = 4, repeats = 3))

mlrFit10 <- train(registered ~ season + yr + holiday + weathersit + atemp + hum + windspeed + I(atemp^2) + I(hum^2)+ I(windspeed^2), data = dayTrain,
    method = "lm",
    preProcess = c("center", "scale"),
    trControl = trainControl(method = "repeatedcv", number = 4, repeats = 3))

mlrFit11 <- train(registered ~ season + yr + holiday + weathersit + atemp + hum + windspeed + I(atemp^2) + I(hum^2) + holiday:weathersit, data = dayTrain,
    method = "lm",
    preProcess = c("center", "scale"),
    trControl = trainControl(method = "repeatedcv", number = 4, repeats = 3))

mlrFit12 <- train(registered ~ season + yr + holiday + weathersit + atemp + hum + windspeed + I(atemp^2) + I(hum^2) + holiday:atemp, data = dayTrain,
    method = "lm",
    preProcess = c("center", "scale"),
    trControl = trainControl(method = "repeatedcv", number = 4, repeats = 3))

comparison <- data.frame(t(mlrFit9$results), t(mlrFit10$results), t(mlrFit11$results), t(mlrFit12$results))
colnames(comparison) <- c("mlrFit9", "mlrFit10", "mlrFit11", "mlrFit12")
comparison
#The lowest RMSE out of the 4 candidate models varies each time I run cross validation, so I will choose the simplest of the 4, mlrFit9
mlrBest <- mlrFit9

# for potentially automating choice of model
# which.min(c(mlrFit9$results["RMSE"], mlrFit10$results["RMSE"], mlrFit11$results["RMSE"], mlrFit12$results["RMSE"]))
```


```{r random_forest, message = FALSE, warning = FALSE}
rfFit <- train(registered ~ . - instant - casual - cnt, data = dayTrain,
               method = "rf",
               trControl = trainControl(method = "repeatedcv", number = 4, repeats = 3),
               preProcess = c("center", "scale"),
               tuneGrid = expand.grid(mtry = c(2, 7, 10:16, 20, 24)))
rfFit
```

I added an initial version of my MLR and random forest models, and wrote a separate R script (ST558RenderProject2.r) to automate the reports for each day of the week.  There's still plenty of clean-up to do with the output, but the automation generally seems to be working.  Since the dataset now includes only one weekday at a time, some of our graphs, tables, etc. that included weekday don't make as much sense.  I still have some questions about whether we're supposed to do any initial data exploration with the full training set, or if we only work with one day at a time.  I might post a question on the discussion board or go to Wednesday office hours unless it's clear to you.